---
title: "world_map_density"
author: "Jefferson Amah"
date: "2025-07-28"
output: html_document
---


## Introduction

This document shows the global population density using R.  
We use geospatial and demographic datasets, primarily from the `rnaturalearth` package.  
The population density (people per km²) is calculated and displayed with color-coded shading.

### Load Required Packages

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(units)   # for area units
```

### Load and Prepare Data

Firstly, we will load the world map with country data

Then calculate the area and convert to km²; we then calculate population density from population estimate and calculated area.

Next, we filter out all null population e.g in Antartica.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

world <- world %>%
     mutate(
         area_km2 = st_area(geometry),                 # get area in m²
         area_km2 = set_units(area_km2, km^2),         # convert to km²
         area_km2 = as.numeric(area_km2),              # drop units to numeric
        pop_density = pop_est / area_km2              # compute density
     )

world_plot <- world %>%
     filter(!is.na(pop_density) & pop_density > 0)
 
```

### Ploting world map

```{r}
ggplot(data = world_plot) +
     geom_sf(aes(fill = pop_density)) +
     scale_fill_viridis_c(trans = "log", option = "C", name = "People per km²") +
     theme_minimal() +
     labs(title = "World Population Density",
          subtitle = "Designed by Geography Class @ Everest School, Ibadan",
          caption = "Source: Natural Earth via rnaturalearth")
```

### Save map to a PNG file

```{r}
ggsave("world_population_density.png", width = 10, height = 6, dpi = 300)
```

### sessionInfo
