---
title: "Africa Population Density Map"
author: "Jefferson Amah"
date: "2025-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(units)
```

## Introduction

This project visualizes the population density of African countries using geographic data from the rnaturalearth package.

Load Africa Data

```{r}
africa <- ne_countries(continent = "Africa", returnclass = "sf")

# Fix invalid geometries
africa <- st_make_valid(africa)

# Project to an equal-area projection (important for correct area in km²)
africa <- st_transform(africa, crs = "+proj=cea +lon_0=25")

# Estimate area in square kilometers
africa <- africa %>%
  mutate(
         area_km2 = st_area(geometry),            # get area in m²
         area_km2 = set_units(area_km2, km^2),    # convert to km²
         area_km2 = as.numeric(area_km2),         # drop units to numeric
        pop_density = pop_est / area_km2          # compute density
     )
```

### Include political structure

``` {r}
# Create centroids (geometric centers) for each country polygon
africa_centroids <- africa %>%
  st_centroid(of_largest_polygon = TRUE) %>%
  st_transform(st_crs(africa)) 
```

### PLotting Africa maps

```{r}
ggplot(data = africa) +
     geom_sf(aes(fill = pop_density)) +
     scale_fill_viridis_c(trans = "log", option = "D", name = "Person(s) per km²", breaks = c(0, 1, 10, 40, 150, 500),
  labels = c("0","1", "10", "100", "1K", "10K")) +
     geom_sf_text(data = africa_centroids, aes(label = name), size = 2.5, color = "black") +
     labs(title = "Population Density of Africa",
          subtitle = "Designed by Geography Class @ Everest School, Ibadan",
          caption = "Source: Natural Earth via rnaturalearth") +
     coord_sf(expand = FALSE) + 
  theme_minimal() +           
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 8),     
    axis.title = element_blank(),         
    plot.caption = element_text(size = 8)
  )
```

### Save map to a PNG file

```{r}
ggsave("africa_population_density.png", width = 10, height = 6, dpi = 300)
```

### Conclusion

This map helps us visualize how population is distributed across Africa. Higher population densities are seen in countries such as Nigeria and Rwanda.
